Experiences with Planning for Natural Language Generation

### 2009-05-30 trunk/Papers/ci-spark/discussion.tex
rewrote discussion

The positive conclusion we draw from our simple set of experiments is that
modern planners are fairly good at controlling the search for many of the
moderately-sized NLG problems we looked at. This is particularly true for
FF in the sentence generation domain, which generates nontrivial 14-word
sentences in about two seconds. Our results also indicate there is still
room for improvement, however: special-purpose algorithms are able to
generate much larger referring expressions in milliseconds
\citep{AreKolStr08}. Although search efficiency is not the main concern in
these domains, it nevertheless becomes a factor in scenarios like
Experiment~2, where the world is almost completely unconstrained and
(unsurprisingly) the search becomes much slower.

The most restrictive bottleneck in the two NLG domains we investigated is
the initial grounding step that many modern planners perform. While this
may be a good strategy for traditional planning benchmarks and IPC
domains---where plan length often dominates universe size and an initial
investment into grounding pays off in saved instantiations during
search---this approach has a very pronounced effect in domains like ours.
As our experiments show, there are natural planning domains in which
relatively short plans must be computed in large universes. For instance,
in the sentence generation domain, it is not unrealistic to look for plans
of length 20 over a universe consisting of thousands of domain individuals
and tens of thousands of possible action instantiations, some of which
require three domain individuals as parameters. Planners like FF and SGPLAN
are less effective in such domains, to the point that grounding
often dominates the total running time of these planners. (In the case of
Experiment~1, our ad-hoc implementation of GraphPlan outperforms both
planners in the sentence generation task.)

Since our domains are not that unusual in their structure and composition,
we hope that the lessons learned from our experiences can help improve the
performance of current planning systems. For instance, the common strategy
of splitting the universe into several types of individuals will not be a
very effective solution in our NLG domains given the large number of
individuals that could exist for a given type. On the other hand,
sophisticated techniques based on reachability analysis may have the
potential to improve this situation, provided such approaches can avoid the
drawbacks of unnecessary grounding.\footnote{See
  \citep{Bryce:07} for a good survey on reachability heuristics for planning
  graph-based approaches to plan generation.}
While recent trends in planning research have (rightfully) focused on the
development of algorithms that control search in sophisticated ways,
resulting in state-of-the-art planners that are more powerful and
successful than their predecessors, we nevertheless believe that
improvements are necessary if planning is to become a more mature
technology that can offer tools to a wider community of users and
researchers. At the end of the day, real-world users will care about the
\emph{total} runtime of a planner, which is more than just the search time.

By and large, our experiences with the planning community from a user's
point of view have been favourable. Thanks to the planning competitions, it
is easy to identify and download a fast state-of-the-art planner, at least
for Linux. However, in the course of our experiments we found and reported
bugs in both FF and SGPLAN. We also discovered that deciding as to which
planner is appropriate for a particular task is not always straightforward,
as the best performing planner in the IPC is not necessarily suitable for
every planning task. As our experiments have shown, even planners as
closely related as FF and SGPLAN can differ significantly in their
performance on different domains (for instance, our results in
Experiment~4).

>>>>

There are both positive and negative conclusions about the state of
planning for modern NLG applications that we can draw from these
experiments.

On the one hand, we found that modern planners are very good at
dealing with the \emph{search} problems that are involved in the
NLG-based planning problems we looked at. In the sentence generation
domain, FF's Enforced Hill-Climbing strategy finds plans corresponding
to 25-word sentences in about a second. It is hard to compare this
number to a baseline because there are no shared benchmark problems;
but FF's search outperforms a greedy search algorithm, and the
performance is competitive with other sentence generators, which
similarly employ special-purpose algorithms. Thus research on search
strategies for planning has paid off: In particular, the Enforced
Hill-Climbing heuristic outperforms the best-first strategy to which
FF 2.3 switches for some problem instances. Similarly, SGPLAN's
performance on the GIVE domain is very convincing and fast enough for
the application.

However, each of the off-the-shelf planners we tested spent
substantial amounts of time on preprocessing. This is most apparent in
the sentence generation domain, where the planners spent almost their
entire runtime on grounding the predicates and operators for some
instances. This effect is much weaker in the GIVE domain, which has a
much smaller number of operators; but as we added dummy predicates,
the ratio of time that the planners spent on preprocessing rose in a
similar way. While the GIVE domain can be carefully defined in such a
way that the number of operators is minimized, this is not possible
for a domain in which the operators encode the different communicative
actions that the NLG system can use. For instance, in the sentence
generation domain, the XTAG planning problem for $k=2$ and $n=5$
consists of about 1000 operators for the different lexicon entries for
all the words in the sentence, some of which take four parameters. It
is not unrealistic to assume a knowledge base with a few hundred
individuals. All this adds up to trillions of ground instances, i.e.\
it is completely infeasible to compute ground instances naively.

This means that while current planners deal well with the search
problems we have looked at, they are still unusable for practical NLG
applications because of the time they spend on preprocessing. In other
words, generation-as-planning research is still not in a much better
position than it was in the 1980s. While we are aware that the time
that a planner invests in preprocessing can pay off during search, we
still suggest that the inability of current planners to scale to
larger domains still limits their usefulness for applications beyond
NLG as well. Furthermore, we feel that preprocessing gets less
research attention than it deserves: If it is scientifically trivial,
we challenge the planning community to come up with implementations
that only ground operators by need; otherwise, we look forward to
publications on this topic. To support this, we offer our planning
domains as benchmarks for future competitions. They are available at
\todo{upload them somewhere}

Finally, we found it very convenient that the recent IPC planning
competitions provide an entry point for selecting and obtaining
current planners. Nevertheless, our experiments exposed several bugs
in the planners we tested, and required us to make changes to the
source code to make them scale to our inputs. We also found that
different planners differ in the variants of PDDL that they can
process. These differences range from fragments of ADL that can be
parsed to sensitivity to the order of declarations and the use of
``objects'' rather than ``individuals'' as the keyword for declaring
the universe. We propose that the case for planning as a mature
technology with professional-quality implementations could be made
more strongly if such discrepancies were harmonized.

### 2009-05-30 trunk/Papers/ci-spark/experiments.tex
rewrote discussion

The results of this experiment are shown in the graph in
Figure~\ref{fig:runtimes-crisp}. The input parameters $(m,n)$ are
plotted in lexicographic order on the horizontal axis, and the runtime
is shown in milliseconds on the vertical axis.\footnote{These times
  were measured on XXX} These results reveal a number of interesting
insights. First, FF significantly outperforms SGPLAN in this domain,
sometimes by a factor of 10 or more.\footnote{Experiments with SGPLAN
  use a pre-release version of SGPLAN, kindly provided to us by
  Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which
  caused it to crash on some of our problem instances.}  Second, FF's
runtime is dominated by its initial grounding step, in which it
computes the ground instances of all predicates and actions used in
the planning problem in order to avoid unnecessary instantiation
during search.  In particular, the ratio of grounding time to total
runtime is generally above 85\%, and rises to above 99\% at $m=11$,
which is still a relatively small universe for this
application.\footnote{The ``grounding'' time reported here is what FF
  reports as ``time spent: instantiating action templates''.}

@@

Fig.~\ref{fig:xtag-graph} reports the runtimes we measured in this
experiment for FF. Because the computations in this experiment were
very memory-intensive, we recompiled FF 2.3 as a 64-bit binary. We
also ran the experiments with a 64-bit version of Metric-FF; although
our planning problem is purely sequential and not metric, Metric-FF
avoids a bug which caused FF 2.3 to switch to a best-first search
strategy beyond a certain point; the jump in runtime is clearly
visible in the figure. We did not run this experiment with SGPLAN
because we didn not have access to a 64-bit version of either SGPLAN
5.2.2 or SGPLAN 6; some initial experiences with the 32-bit version
indicate that SGPLAN switches to FF's best-first strategy even for
small problem instances, and thus would be slower than the versions of
FF we reported here. We also do not report runtimes for our Java
implementation of Graphplan, because it was unusuably slow for serious
problem instances: For $k=1$ and $n=3$, it already took over two
minutes, and it exceeded its memory limit of 16 GB for
$n>3$.

There are a number of observations we can make in this
experiment. First of all, the experiment confirms that FF's Enforced
Hill-Climbing search strategy works very well for the sentence
generation task: Although we are now generating with a large grammar,
FF generates a 39-word sentence ($k=1,n=6$) in under a second of
search time. This efficiency is a direct result of this particular
search strategy: Whenever FF falls back to the best-first strategy,
runtimes become much slower (compare the graphs for FF 2.3 and
Metric-FF). Second, FF's runtime is still dominated by the
preprocessing stage. For instance, Metric-FF spends about 10 seconds
on search for $k=1,n=10$, compared to its total runtime of about 65
seconds. This effect becomes more pronounced as we increase $k$: For
$k=2$, we reach 65 seconds of total runtime at $n=4$, but here
Metric-FF only spends about a second on search. For $k=3$, \todo{say
  something true}. This is to be expected, given that the planning
operators for the verbs have $k+2$ parameters (see
Fig.~\ref{fig:white-rabbit-as-planning}), and thus the number of
action instances grows by a factor of the universe size every time we
increase $k$ by one. A planner which computes all ground instances of
the operators thus takes exponential time in $k$ for preprocessing.

>>>>

The results of this experiment are shown in the graph in
Figure~\ref{fig:runtimes-crisp}.\footnote{All
  runtimes were measured on a Pentium 4 CPU running at 3 GHz. Java programs
  were allowed to ``warm up'', i.e., the planner was run five times and the
  first four measurements discarded to ensure that the JVM had just-in-time
  compiled all relevant bytecode.} The input parameters $(m,n)$ are
plotted in lexicographic order on the horizontal axis, and the runtime
is shown in milliseconds on the vertical axis.\footnote{These times
  were measured on XXX} These results reveal a number of interesting
insights. First, FF significantly outperforms SGPLAN in this domain,
sometimes by a factor of 10 or more.\footnote{Experiments with SGPLAN
  use a pre-release version of SGPLAN, kindly provided to us by
  Chih-Wei Hsu.The release version of SGPLAN 5.2.2 had a bug which
  caused it to crash on some of our problem instances.}  Second, FF's
runtime is dominated by its initial grounding step, in which it
computes the ground instances of all predicates and actions used in
the planning problem in order to avoid unnecessary instantiation
during search.  In particular, the ratio of grounding time to total
runtime is generally above 85\%, and rises to above 99\% at $m=11$,
which is still a relatively small universe for this
application.\footnote{The ``grounding'' time reported here is what FF
  reports as ``time spent: instantiating action templates''.}

@@

Fig.~\ref{fig:xtag-graph} reports the runtimes we measured in this
experiment for FF. Because the computations in this experiment were
very memory-intensive, we recompiled FF 2.3 as a 64-bit binary. We
also ran the experiments with a 64-bit version of Metric-FF; although
our planning problem is purely sequential and not metric, Metric-FF
avoids a bug which caused FF 2.3 to switch to a best-first search
strategy beyond a certain point; the jump in runtime is clearly
visible in the figure. We did not run this experiment with SGPLAN
because we didn not have access to a 64-bit version of either SGPLAN
5.2.2 or SGPLAN 6; some initial experiences with the 32-bit version
indicate that SGPLAN switches to FF's best-first strategy even for
small problem instances, and thus would be slower than the versions of
FF we reported here. We also do not report runtimes for our Java
implementation of Graphplan, because it was unusuably slow for serious
problem instances: For $k=1$ and $n=3$, it already took over two
minutes, and it exceeded its memory limit of 16 GB for
$n>3$. \todo{compare to SPUD reconstruction here}

There are a number of observations we can make in this
experiment. First of all, the experiment confirms that FF's Enforced
Hill-Climbing search strategy works very well for the sentence
generation task: Although we are now generating with a large grammar,
FF generates a 39-word sentence ($k=1,n=6$) in under a second of
search time. This efficiency is a direct result of this particular
search strategy: Whenever FF falls back to the best-first strategy,
runtimes become much slower (compare the graphs for FF 2.3 and
Metric-FF). Second, FF's runtime is still dominated by the
preprocessing stage. For instance, Metric-FF spends about 10 seconds
on search for $k=1,n=10$, compared to its total runtime of about 65
seconds. This effect becomes more pronounced as we increase $k$: For
$k=2$, we reach 65 seconds of total runtime at $n=4$, but here
Metric-FF only spends about a second on search. For $k=3$, neither FF
nor Metric-FF were able to solve any of the input
instances. \todo{bla} This is to be expected, given that the planning
operators for the verbs have $k+2$ parameters (see
Fig.~\ref{fig:white-rabbit-as-planning}), and thus the number of
action instances grows by a factor of the universe size every time we
increase $k$ by one. A planner which computes all ground instances of
the operators thus takes exponential time in $k$ for preprocessing.
